---
title: "ED Assignment"
format: pdf
editor: visual
author: Malik Jamie (JMXMAL003) and Amara Patel (PTLAMA005)
date: 21/10/2025
---

# Introduction

The aim of this experiment is to examine the effects of different intensities of heat and light on the growth of chilli plants. The goal is to determine whether light and heat affect chilli growth, which settings produce the highest quality chillies and by how much. Furthermore, whether these settings depend on variety as well as whether quality depends on variety will be explored. This is to make recommendations about which heat and light settings are optimal for small-scale farmers to implement, taking into account the cost of additional heat and light.

These goals will be tackled through an analysis of variance of this exploratory experiment.

## Section 1: Design and Randomisation

experimental unit: plots (only one treatment per plot); 32 plots over two greenhouses blocking factor 1: north/south blocking factor 2: greenhouse (A, B) blocking factor 3: season (1, 2)

design idea: overarching block: season 1 and season 2 (randomise the 32 treatments to the 32 plots (16 treatments per greenhouse) blocking with season for greenhouse and side of the greenhouse so possible latin square design/nest the blocks)

then we can remove one of these two blocking factors above (ie overarching block: season; each treatment of heat and light (16 total) randomised to 32 plots (one in each greenhouse), with one of each type of chilli in each plot; then we block for north/south - in this case, experimental unit would be plants.

hypotheses (i honestly don't know how i should be defining these... possibly using contrasts but not sure. also not sure how i should be setting up the model but i am busy doing more stats and reading to figure it all out): 1: Does light have an effect on the quality of chillies $H_0: \alpha_{light} = 0$ $H_1: \alpha_{light} \ne 0$

nah nah nah

2: Which light setting has the biggest effect 2: Does heat have an effect on the quality of chillies

the code chunk below is basically just setting up a possible design where i've randomised variety over greenhouse-season blocks and then randomised light and heat within each greenhouse. not sure what this means for the analysis of the effects of variety (do experimental units for variety change from plot to greenhouse, w two reps or can we just analyse the 64 individual plots and make conclusions about those while adjusting for the greenhouse-season blocking factor?) i honestly don't know how this project is supposed to be approached and i feel a little bit lost to be honest. i might go to the lab session on tuesday and ask some of the tutors for advice.


```{r another model based on what ethan told me}
library(dplyr)
library(knitr)
library(tibble)
library(agricolae)
library(tidyr)
library(EDproject)
library(ggplot2)

# model: essentially, side and variety are treated as treatment factors w 2 levels each, resulting in 4 combinations in a factorial experiment. light and heat are treated as blocking factors within each greenhouse-season block. 4, 4x4 lsds are created with rows corresponding to light levels and columns corresponding to heat levels and the four side-variety treatments are assigned in a standard 4x4 lsd. the model ends up being Y = mu + heat + light + season + greenhouse + side + variety + side*variety + heat*light + variety*heat + variety*light + variety*heat*light + error term. end up with 16 replicates per treatment of side-variety

set.seed(15)

# treatments 
treatments2 <- c("n,R", "s,F", "n,F", "s,R")

# create design (again oh my god)

mydesign2 <- expand.grid(season = 1:2, greenhouse = c("A", "B") ) %>%
  
  # new lsd for each season-greenhouse block w individual seed
  mutate(
    square_id = 1:n(),
    latin_seed = 15 + square_id
  ) %>%
  rowwise() %>% 
  mutate(
    latin_square = list(
      design.lsd(treatments2, serie = 0, seed = latin_seed)$book)
  ) %>%
  ungroup() %>% 
  
  # getting latin square into separate elements to be able to put it into the data frame 
  tidyr::unnest(latin_square) %>%
  rename(
    row = row,
    column = col,
    treatment = treatments2
  ) %>%
  
  # splitting the treatments into individual factors applied to put into design
  separate(treatment, into = c("side", "variety"), sep = ",") %>%
mutate(
    light = row,    # rows correspond to light levels (1-4)
    heat = column   # columns correspond to heat levels (1-4)
  ) %>%
  
# adding plot numbers (1-32 for season 1, 2)
  group_by(season) %>%
  mutate(plot = 1:n()) %>%
  ungroup() %>%
  select(plot, season, greenhouse, side, variety, light, heat) %>%
  arrange(season, plot) 

write.csv(mydesign2, file = "design2.csv", row.names = FALSE)

```

```{r}

mydesign <- read.csv("design.csv")
mydata <- get.observations(mydesign)

mydesign2 <- read.csv("design2.csv")
mydesign2

mydata2 <- get.observations(mydesign2)

```
