---
title: "STA3041F 2026 Tutorial Assessment 4"
author: "Amara Patel (PTLAMA005) and Malik Jamie (JMXMAL003)"
date: '2026-04-10'
output: pdf_document
header-includes:
  - \usepackage{pdfpages}
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,fig.align='center',fig.width=5,fig.height=5)
```

## Question 1


### (a)
The probability of 0 offspring in a generation is directly dependent on time (i.e. the generation m). In this time dependence, the un-normalised probability (before dividing by the total of all the entries to get actual probabilities that sum to 1) always stays above or equal to 3 and oscillates since it's dependent on a sine function. Since the expression (1 + 0.03(1+sin(m*pi/5))) is always greater than one and raised to the power of m, as m increases, the probability is going to increase too, even though it will oscillate between 1 and 1.06. This means that as m gets larger, the probability of having 0 children will also get exponentially larger which, after normalising, will result in the probabilities of having 1, 2 or 3 children tend to 0 since they are not exponentially increasing with time. This means that intuitively, the branching process will almost certainly go extinct as m increases causing the probability of having 0 children to get exponentially bigger, despite the generation dependent oscillation factor. 

### (b)
```{r }
sim = function() # making a function
{
	 Z = 1 # setting beginning population size
	 k = 1 # setting generation to start 
   while(Z!=0)
   {
	  X_states = c(0, 1, 2, 3) # number of children
	prob_X   = c(3*(1+0.03*(1+sin((k*pi)/5)))^k, 2, 4, 1) # unnormalised probabilities 
	  prob_X   = prob_X/sum(prob_X) # normalise
	  Z        = sum(sample(X_states, Z, replace = TRUE, prob = prob_X)) # sample to get population at kth generation 
	  k        = k + 1 # iterate 
   }
	 if(Z==0) # extinction event 
	 {
	 	 return(k-1) # time before extinction event occurs
	 }else{return(NA)}
}

# simulation to get empirical distribution of first passage 
set.seed(8)
sims <- replicate(10000, sim())

hist(sims, xlab = "first passage time to extinction", main = "Histogram of first passage time to extinction", breaks = seq(min(sims)-0.5, max(sims)+0.5, by=1), freq = FALSE, col = "pink2")
```

### (c)
\begin{equation*}
\begin{aligned}
Eqn here.
\end{aligned}
\end{equation*}

### (d)


```{r }
Gx = function(s,m)
{

}

Gm = function(s,m)
{
	
}

```






\newpage
\includepdf[pages=1]{Department of Statistical Sciences - Plagiarism Declaration.pdf}


