---
title: "STA3041F 2026 Tutorial Assessment 4"
author: "Amara Patel (PTLAMA005) and Malik Jamie (JMXMAL003)"
date: '2026-04-10'
output: pdf_document
header-includes:
  - \usepackage{pdfpages}
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,fig.align='center',fig.width=5,fig.height=5)
```

## Question 1


### (a)
The probability of 0 offspring in a generation is directly dependent on time (i.e. the generation m). In this time dependence, the un-normalised probability (before dividing by the total of all the entries to get actual probabilities that sum to 1) always stays above or equal to 3 and oscillates since it's dependent on a sine function. Since the expression (1 + 0.03(1+sin(m*pi/5))) is always greater than one and raised to the power of m, as m increases, the probability is going to increase too, even though it will oscillate between 1 and 1.06. This means that as m gets larger, the probability of having 0 children will also get exponentially larger which, after normalising, will result in the probabilities of having 1, 2 or 3 children tend to 0 since they are not exponentially increasing with time. This means that intuitively, the branching process will almost certainly go extinct as m increases causing the probability of having 0 children to get exponentially bigger, despite the generation dependent oscillation factor. 

### (b)
```{r }
sim = function() # making a function
{
	 Z = 1 # setting beginning population size
	 k = 1 # setting generation to start 
   while(Z!=0)
   {
	  X_states = c(0, 1, 2, 3) # number of children
	prob_X   = c(3*(1+0.03*(1+sin((k*pi)/5)))^k, 2, 4, 1) # unnormalised probabilities 
	  prob_X   = prob_X/sum(prob_X) # normalise
	  Z        = sum(sample(X_states, Z, replace = TRUE, prob = prob_X)) # sample to get population at kth generation 
	  k        = k + 1 # iterate 
   }
	 if(Z==0) # extinction event 
	 {
	 	 return(k-1) # time before extinction event occurs
	 }else{return(NA)}
}

# simulation to get empirical distribution of first passage 
set.seed(8)
sims <- replicate(10000, sim())

hist(sims, xlab = "first passage time to extinction", main = "Histogram of first passage time to extinction", breaks = seq(min(sims)-0.5, max(sims)+0.5, by=1), freq = FALSE, col = "cyan4")
```

### (c)

For a time-inhomogeneous branching process, the generating function of changes depending on the generation m, and so to derive the generating function, the generation specific generating functions must be composed sequentially up to the (m-1). Evaluating from the outside in, the expression for the PGF of $Z_m$ is:$$G_m(s) = G_X(G_X(\dots G_X(s, m-1) \dots, 2), 1)$$

### (d)


```{r q4}
Gx = function(s,m)
{
  # probabilities
  p0 <- 3*(1+0.03*(1+sin((k*pi)/5)))^m
  p1 <- 2
  p2 <- 4
  p3 <- 1
  # normalise probabilities 
  p0 <- p0/sum(c(p0, p1, p2, p3))
  p1 <- p1/sum(c(p0, p1, p2, p3))
  p2 <- p2/sum(c(p0, p1, p2, p3))
  p3 <- p3/sum(c(p0, p1, p2, p3))
  
  # calculate Gx = E(s^X)
  Gx <- p0 + p1 * s + p2 * s^2 + p3 *s^3 
  return(Gx)

}

Gm = function(s,m)
{ 
  if (m==1) {
    return(s)
  }
  Gm <- s
  # compose functions from inside out so iterate from m-1 to 1 
  for(k in m-1:1) {
  Gm <- Gx(Gm, k)
}
	return(Gm)
}

# find distribution of first passage (use G_m(0)-G_m-1(0))

## the histogram again
set.seed(8)
sims <- replicate(10000, sim())

hist(sims, xlab = "first passage time to extinction", main = "Histogram of first passage time to extinction", breaks = seq(min(sims)-0.5, max(sims)+0.5, by=1), freq = FALSE, col = "lightpink")

## find the true probabilities to superimpose 
max_k <- max(sims, na.rm = TRUE)
true_times <- 1:max_k
true_probs <- numeric(max_k)

for (k in true_times){
  true_probs[k] <- Gm(0, k+1) - Gm(0, k)
}

lines(true_times, true_probs, type = "p", pch = 16, col = "purple4")
```









