---
title: "STA3041F 2026 Tutorial Assessment 3"
author: "Amara Patel (PTLAMA005) and Malik Jamie (JMXMAL003)"
date: '2026-03-27'
output: pdf_document
header-includes:
  - \usepackage{pdfpages}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(stats)
```

## Question 1

```{r q1, fig.align='center',fig.width=5,fig.height=5, fig.cap='Figure caption here.'}
rm(list = ls())
dat = read.table("BatRun Data 2.txt", h = TRUE)

# Plot the count trajectory and place tick marks for the arrivals
plot(dat$Pos~dat$Sn, type='s', xlab = "Time in minutes from 03:30 hour mark", ylab = "Count", main = "Count Trajectory of BATRUN Race")
rug(dat$Sn, side=1, col="pink")

```

## Question 2

\begin{equation*}
\begin{aligned}
\mbox{mvf}(v;\boldsymbol\beta) &= \int_0^v \lambda(u;\boldsymbol\beta)du\\
 &= \int_0^v \exp\left(-\beta_1^2\beta_2 + 2\beta_1\beta_2u - \beta_2u^2 \right)du\\
 &=\int_0^v \exp\left(-\beta_2 (u - \beta_1)^2 \right) du \\
 &\text{let } s = \sqrt{2\beta_2}(u-\beta_1), \text{ then } du = \frac{1}{\sqrt{2\beta_2}} ds \\
 &= \int_{\sqrt{2\beta_2}(-\beta_1)}^{\sqrt{2\beta_2}(v-\beta_1)} \exp\left( -\frac{s^2}{2} \right) \frac{1}{\sqrt{2\beta_2}} ds \\
 &= \sqrt{\frac{\pi}{\beta_2}} \int_{\sqrt{2\beta_2}(-\beta_1)}^{\sqrt{2\beta_2}(v-\beta_1)} \frac{1}{\sqrt{2\pi}} \exp\left( -\frac{s^2}{2} \right) ds \\
 &= \sqrt{\frac{\pi}{\beta_2}} \left[ \Phi\Big(\sqrt{2\beta_2}(v-\beta_1)\Big) - \Phi\Big(\sqrt{2\beta_2}(-\beta_1)\Big) \right] \\
&= K(\beta_2) \times \Phi\left(u; \beta_1, \frac{1}{\sqrt{2\beta_2}}\right) \text{ where } K(\beta_2) = \sqrt{\frac{\pi}{\beta_2}}
\end{aligned}
\end{equation*}

## Question 3

```{r q3}

# for fixed n (n=50)
mvf_1    = function(v,pars)
{
		beta = pars
		mvf = ((1/(beta[2])) * exp(beta[1])) * (exp(beta[2]*v) - 1)
	  return(mvf)
}

min_likelihood_1 = function(pars)
{
	  beta  = pars
	  s <- dat$Sn
	  log_intensity = beta[1]+beta[2]*s
	  value =  sum(log_intensity) - mvf_1(max(s), beta)
	  return(-value)
}

mvf_2    = function(v,pars)
{
    beta = pars
	  K     =  sqrt(pi/beta[2])
# limits 
	  upper = (v-beta[1]) * sqrt(2*beta[2])
	  lower = -beta[1] * sqrt(2*beta[2])
	  mvf = K * (pnorm(upper) - pnorm(lower))
	  
	  return(mvf)
}

min_likelihood_2 = function(pars)
{

  beta = pars
	s <- dat$Sn
  log_intensity = -(beta[1]^2*beta[2]) + (2*beta[1]*beta[2]*s) -(beta[2]*s^2)
	  value =  sum(log_intensity) - mvf_2(max(s),pars)
	  return(-value)
}

```

## Question 4

```{r q4, warning=FALSE}
set.seed(147)
p <- c(mean(dat$Sn), runif(1))

res_opt_1 = nlm(min_likelihood_1, p)
res_opt_2 = nlm(min_likelihood_2, p)

AIC_1 = 2*2 + 2*res_opt_1$minimum
AIC_2 = 2*2 + 2*res_opt_2$minimum
  
M1  = c(res_opt_1$estimate, AIC_1)
M2  = c(res_opt_2$estimate, AIC_2)
tab = cbind(M1,M2)
rownames(tab) = c('beta[1]','beta[2]','AIC')
colnames(tab) = c('Model 1','Model 2')
tab
```

## Question 5

```{r q5, fig.align='center',fig.width=5,fig.height=5, warning=FALSE}

set.seed(147)
p <- c(mean(dat$Sn), runif(1))
t_seq <- seq(0, max(dat$Sn), length.out=50) #to smoothe out mean trajectory

res_opt_1 = nlm(min_likelihood_1, p)
res_opt_2 = nlm(min_likelihood_2, p)

mle_1 = res_opt_1$estimate
mle_2 = res_opt_2$estimate

#tt = 
E1 = mvf_1(t_seq, mle_1)
E2 = mvf_2(t_seq, mle_2)

plot(dat$Pos~dat$Sn, type='s', xlab = "Time in minutes from 03:30 hour mark", ylab = "Count", main = "Count Trajectory of BATRUN Race")
lines(t_seq, E1, col="light blue")
lines(t_seq, E2, col="red3")

```
Based on the model statistics calculated in Question 4, Model 1 yielded an extremely high AIC (2.531127e+33), suggesting that it fits very poorly to the observed data. Model 2 yielded a much lower AIC (157.7273), which is far more appropriate given the observed data and indicates a fairly good trade-off between model complexity and goodness-of-fit.
Furthermore, Model 2 yielded a large negative estimate for $\beta_2 = -4.929 \times 10^{1}$. Since Beta 2 represents the growth rate of the rate of arrivals, this estimate suggests that the arrival rate drops off dramatically for t>0. However, the plot of the observed data shows that the arrival rate of runners actually increased as t got larger. This again demonstrates the poor fit that Model 1 provides for the data.
In contrast, Model 2 yielded a positive estimate for $\beta_2 = 2.287\times 10^{-4}$, which appears to be appropriate for the observed data.
For Question 5, we were actually unable to overlay the mean trajectory for Model 1, due to how poor it fit the observed data.
The mean trajectory for Model 2 fits the observed trajectory very well, suggesting that Model 2 is a good fit for the observed data.
For the above reasons, we would strongly prefer using Model 2 to make predictions for this data.


## Question 6

```{r q6}
forecast = mvf_2(60, mle_2) - mvf_1(0, mle_2)
forecast
```
Model 2 predicts that between 10 and 11 runners will finish between 03:30:00 and 04:30:00 for the 2026 edition of the BATRUN. 11 runners finished between these times in the 2025 edition of the BATRUN, so this predicted value appears sensible.
